---
title: "BHM Basket Trial Logistic Regression - RStan v JAGS v Nimble"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Two-arm trial TFP v RStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
#on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

# If on GitHub, set eval = FALSE for all subsequent chunks
#knitr::opts_chunk$set(eval = !on_github)

```

## Quick start
This vignette compares the same Bayesian hierarchical model fitted with Rstan, JAGS and Nimble, and is similar the previous vignette for a logistic regression but extened to a basket trial design scenario with two baskets. Only key snippets and outputs are shown below. The full vignette is available at [here](https://raw.githubusercontent.com/GSK-Biostatistics/vignettes-asia/refs/heads/main/jags_nimble/stan_jags_nimble2.qmd.qmd). 

### Model Formulation

The model implemented here is a non-centered parameterization for a Basket trial with two baskets and a logistic response model.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\tilde{\theta}_j &\sim \text{Normal}(0, 1)\quad\enspace\text{for}\enspace{j=1,\dots,4}  \\
\beta_{0,0} &= \mu_0 + \sigma_0 \tilde{\theta}_1 \quad\quad\hspace{0.2cm}\text{ intercept - log odds control basket 0}\\
\beta_{0,1} &= \mu_0 + \sigma_0 \tilde{\theta}_2 \quad\quad\hspace{0.2cm}\text{ intercept - log odds control basket 1}\\
\beta_{1,0} &= \mu_1 + \sigma_1 \tilde{\theta}_3 \quad\quad\hspace{0.2cm}\text{ treatment - log OR basket 0}\\
\beta_{1,1} &= \mu_1 + \sigma_1 \tilde{\theta}_4 \quad\quad\hspace{0.2cm}\text{ treatment - log OR basket 1}\\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$


```{r rsetup, include=FALSE,eval=TRUE}
library(rstan)
library(rjags)
library(coda)
library(nimble)
library(nimbleHMC)
library(nimbleSCR)
library(bayesplot)
library(ggplot2)
library(kableExtra)
```

## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a (dummy) basket ID variable (=1)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=TRUE,eval=TRUE}
### Prepare model inputs
set.seed(99999)
# Set up data
rr_k_ctrl <- c(0.20,0.25)        # control response rate for each basket
rr_k_trt <- c(0.40,0.45)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(50, K)     # number of control participants per basket
N_k_trt <- rep(50, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

z_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  z_vec<-c(z_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=z_vec)


```


```{r datasetsQ, include=TRUE,eval=TRUE}
my_table1_object <- head(thedata) %>%
  kable() %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = F)#, color = "blue")

my_table1_object

```

## RStan

The code for implementing the above model is RStan is given below. 

```{r stanmodeldef,include=TRUE,echo=TRUE,eval=TRUE}
### Stan model
  BHM_stan_1 <- "data {
  int<lower = 0> K;                     // number of baskets (K >= 2)
  int<lower = 0> N;                     // total number of participants
  int<lower = 0, upper = N> N_k[K];     // K x 1 vector of basket sample sizes
  //int<lower = 0, upper = N> N_k;
  int<lower = 1, upper = K> k_vec[N];   // N x 1 vector of basket indicators
  int<lower = 0, upper = 1> z_vec[N];   // N x 1 vector of treatment indicators for active treatment arm
  int<lower = 0, upper = 1> y[N];       // N x 1 vector of binary responses
}
parameters {
  matrix[K,2] beta_tr;                  // K x 2 matrix of transformed regression coefficients
  real mu1;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma1;                  // scalar of hierarchical SD (log odds ratio)
  real mu0;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma0;                  // scalar of hierarchical SD (log odds ratio)

}
transformed parameters {
  matrix[K,2] beta;                     // K x 2 matrix of regression coefficients (without transformation)
  // Calculate beta coefs using transformed betas (leads to better mixing)
  for (k in 1:K){
    beta[k,1] = mu0 + sigma0 * beta_tr[k,1];
    beta[k,2] = mu1 + sigma1 * beta_tr[k,2];
  }

}
model {
  mu0 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma0 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  //sigma0 ~ exponential(1);
  //sigma0 ~ cauchy(0, 5);
  //sigma0 ~ gamma(2, 0.1);
  mu1 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma1 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  //sigma1 ~ exponential(1);
  //sigma1 ~ cauchy(0, 5);
  //sigma1 ~ gamma(2, 0.1);
  beta_tr[,1] ~ normal(0, 1);  // vectorized normal prior for the log odds for the control arm
  beta_tr[,2] ~ normal(0, 1);  // vectorized normal prior for the log odds ratios
  for (i in 1:N)
    y[i] ~ bernoulli_logit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i]);
}
"

```



```{r stansample,echo=TRUE,include=FALSE,results='hide',eval=TRUE}

# Create list with input values for Stan model
data_input_1 <- list(
  K = K,                # number of subgroups
  N = N,                # total sample size
  N_k = N_k,            # sample size per basket
  k_vec = k_vec,        # N x 1 vector of subgroup indicators
  z_vec = z_vec,        # N x 1 vector of active treatment arm indicators
  y = y                 # N x 1 vector of binary responses
  )



### Compile and fit model
options(mc.cores = parallel::detectCores())
# Compile model (only run the line below once for a simulation study - compilation not dependent on any
# simulation inputs as defined by scenarios or on simulated data)
stan_mod_1 <- stan_model(model_code = BHM_stan_1)
# Fit model
nsamps <- 30000       # number of posterior samples (after removing burn-in) per chain
nburnin <- 1000       # number of burn-in samples to remove at beginning of each chain
nchains <- 4          # number of chains
BHM_pars_1 <- c("mu0", "sigma0","mu1", "sigma1", "beta","beta_tr")     # parameters to sample
start_time <- Sys.time()
stan_fit_2 <- sampling(stan_mod_1, data = data_input_1, pars = BHM_pars_1,
                       iter = nsamps + nburnin, warmup = nburnin, chains = nchains)
end_time <- Sys.time()
end_time - start_time
post_draws_2 <- as.matrix(stan_fit_2)           # posterior samples of each parameter

beta_stan <- extract(stan_fit_2,"beta[1,1]")$"beta[1,1]"

```

### Diagnostics
```{r, standiag, eval=TRUE, include=FALSE}
# Log posterior
lp_fit_2 <- log_posterior(stan_fit_2)
head(lp_fit_2)

#Nuts patterns
np_fit_2 <- nuts_params(stan_fit_2)
head(np_fit_2)

#mcmc parcoord
color_scheme_set("darkgray")
posterior_fit_2 <- as.array(stan_fit_2)
mcmc_parcoord(posterior_fit_2, np = np_fit_2)

fit_summary <- summary(stan_fit_2)$summary

# View the Rhat column for all parameters
print(fit_summary[, c("mean", "Rhat")])

```

### Summarize results

```{r stanresults,fig.width=10, fig.height=6, out.width="100%",eval=TRUE}
library(bayesplot)
color_scheme_set("viridisA")
p<-mcmc_trace(stan_fit_2,"mu0")
#ggsave("precomputed/vg2_traceplot1_stan.png", p)
print(p)

plot(stan_fit_2)
```


### JAGS

```{r, jagsmod, include=FALSE,results='hide',eval=TRUE}
BHM_jags_2 <- "
model {
for (i in 1:N) {
    p[i] <- ilogit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i])
    y[i] ~ dbern(p[i])
  }

  for (k in 1:K) {
    beta_tr[k, 1] ~ dnorm(0, 1)
    beta_tr[k, 2] ~ dnorm(0, 1)
  }
  
  for (j in 1:K){
    beta[j,1] <- mu0 + sigma0 * beta_tr[j,1]
    beta[j,2] <- mu1 + sigma1 * beta_tr[j,2]
  }

  #tau0 <- 1 / (sigma0^2)
  #tau1 <- 1 / (sigma1^2)

  tau_mu <- 1 / (2.5^2)

  mu0 ~ dnorm(0, tau_mu)
  mu1 ~ dnorm(0, tau_mu)

  sigma0 ~ dunif(0, 3)
  sigma1 ~ dunif(0, 3)
}
"

bhm_jags_mod <- jags.model(
  textConnection(BHM_jags_2),
  data = data_input_1,
  n.chains = 4,
  n.adapt = 10000
)

update(bhm_jags_mod, 10000)

jags_mod_sim = coda.samples(model = bhm_jags_mod,
                       variable.names = c("mu0", "sigma0","mu1", "sigma1", "beta","beta_tr"),
                       n.iter = 10000, thin =1)

summary(jags_mod_sim)

#plot(jags_mod_sim)

mat_jags <- as.matrix(jags_mod_sim)
beta_jags <- mat_jags[, "beta[1,1]"]
```


### NIMBLE with default sampler

```{r, nimblemod, include=TRUE, results='hide',eval=TRUE}
nimble_code_1 <- nimbleCode({  

  for (i in 1:N) {
    p[i] <- ilogit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i])
    y[i] ~ dbern(p[i])
  }

  for (k in 1:K) {
    beta_tr[k, 1] ~ dnorm(0, 1)
    beta_tr[k, 2] ~ dnorm(0, 1)
  }
  
  for (j in 1:K){
    beta[j,1] <- mu0 + sigma0 * beta_tr[j,1]
    beta[j,2] <- mu1 + sigma1 * beta_tr[j,2]
  }

  #tau0 <- 1 / (sigma0^2)
  #tau1 <- 1 / (sigma1^2)

  tau_mu <- 1 / (2.5^2)

  mu0 ~ dnorm(0, tau_mu)
  mu1 ~ dnorm(0, tau_mu)

  sigma0 ~ dunif(0, 3)
  sigma1 ~ dunif(0, 3)
  }

) 

#Initial values

K_val <- data_input_1$K
N_val <- data_input_1$N

inits <- list(
  mu0 = 0,
  mu1 = 0,
  sigma0 = 1,
  sigma1 = 1,
  beta_tr = matrix(0, nrow = K_val, ncol = 2),
  beta = matrix(0, nrow = K_val, ncol = 2)
)

# Build model


model <- nimbleModel(
  nimble_code_1,
  # Ensure k_vec is in constants, not data
  constants = list(
    N = N_val,
    K = K_val,
    k_vec = data_input_1$k_vec,
    z_vec = data_input_1$z_vec
  ),
  data = list(
    y = data_input_1$y
  ),
  inits = inits,
  dimensions = list(
    beta_tr = c(K_val, 2),
    beta = c(K_val, 2),
    k_vec = c(N_val)    
  )
)

#Configure the default sampler
# NIMBLE automatically detects conjugacy (Gibbs) and assigns 
# Adaptive Metropolis to non-conjugate parameters.
mcmcConf <- configureMCMC(model)
mcmcConf$addMonitors("beta")
#conf$addMonitors(c("mu0", "mu1", "sigma0", "sigma1", "beta_tr", "beta"))



#Build MCMC
mcmc <- buildMCMC(mcmcConf)

# Compile
Cmodel <- compileNimble(model)

# Build and compile HMC sampler
Cmcmc <- compileNimble(mcmc, project = model)

# Run the sampler

samples <- runMCMC(
  Cmcmc,
  niter = 20000,
  nburnin = 5000,
  thin = 10,
  nchains = 4,
  samplesAsCodaMCMC = TRUE
)

# Diagnosis

library(coda)

#plot(samples)
summary(samples)
gelman.diag(samples)
effectiveSize(samples)

all_samples <- do.call(rbind, samples)

beta_nimble1 <- all_samples[, "beta[1, 1]"]
```


### NIMBLE with HMC

```{r, nimblehmc, include=TRUE, results='hide',eval=TRUE}
nimble_code_1 <- nimbleCode({  

  for (i in 1:N) {
    p[i] <- ilogit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i])
    y[i] ~ dbern(p[i])
  }

  for (k in 1:K) {
    beta_tr[k, 1] ~ dnorm(0, 1)
    beta_tr[k, 2] ~ dnorm(0, 1)
  }
  
  for (j in 1:K){
    beta[j,1] <- mu0 + sigma0 * beta_tr[j,1]
    beta[j,2] <- mu1 + sigma1 * beta_tr[j,2]
  }

  #tau0 <- 1 / (sigma0^2)
  #tau1 <- 1 / (sigma1^2)

  tau_mu <- 1 / (2.5^2)

  mu0 ~ dnorm(0, tau_mu)
  mu1 ~ dnorm(0, tau_mu)

  sigma0 ~ dunif(0, 3)
  sigma1 ~ dunif(0, 3)
  }

) 

#Initial values

K_val <- data_input_1$K
N_val <- data_input_1$N

inits <- list(
  mu0 = 0,
  mu1 = 0,
  sigma0 = 1,
  sigma1 = 1,
  beta_tr = matrix(0, nrow = K_val, ncol = 2),
  beta = matrix(0, nrow = K_val, ncol = 2)
)

# Build model


model <- nimbleModel(
  nimble_code_1,
  # Ensure k_vec is in constants, not data
  constants = list(
    N = N_val,
    K = K_val,
    k_vec = data_input_1$k_vec,
    z_vec = data_input_1$z_vec
  ),
  data = list(
    y = data_input_1$y
  ),
  inits = inits,
  dimensions = list(
    beta_tr = c(K_val, 2),
    beta = c(K_val, 2),
    k_vec = c(N_val)    
  ),
  buildDerivs = TRUE
)

# Compile
Cmodel <- compileNimble(model)

# Key step: Configure HMC

conf <- configureHMC(model)
conf$addMonitors(c("mu0", "mu1", "sigma0", "sigma1", "beta_tr", "beta"))
conf$printSamplers()

# Build and compile HMC sampler
Rmcmc <- buildMCMC(conf)
Cmcmc <- compileNimble(Rmcmc, project = model)

# Run the sampler

samples <- runMCMC(
  Cmcmc,
  niter = 20000,
  nburnin = 5000,
  thin = 10,
  nchains = 4,
  samplesAsCodaMCMC = TRUE
)

# Diagnosis

library(coda)

#plot(samples)
summary(samples)
gelman.diag(samples)
effectiveSize(samples)

all_samples <- do.call(rbind, samples)

beta_nimble2 <- all_samples[, "beta[1, 1]"]

```

### Plot and compare

```{r, plot_res, plot_res, eval=TRUE, include=TRUE}
df <- data.frame(
  value = c(beta_stan, beta_jags, beta_nimble1, beta_nimble2),
  source = factor(rep(
    c("Stan", "JAGS", "NIMBLE", "NIMBLE - HMC"),
    times = c(length(beta_stan), length(beta_jags), length(beta_nimble1), length(beta_nimble2))
  ))
)

library(ggplot2)

ggplot(df, aes(x = value, color = source)) +
  geom_density(size = 1) +
  labs(title = "Posterior density of beta[1,1]",
       x = "beta[1,1]",
       y = "Density") +
  theme_minimal()
  
  


```
