---
title: "EXNEX Basket Trial Logistic Regression using RStan"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Two-arm trial TFP v RStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

###Setup

```{r include=FALSE,eval=TRUE}
library(rstan)
library(dplyr)
library(ggplot2)
library(bayesplot)
```

### Data preparation

```{r dataset, include=TRUE,eval=TRUE}

## trying six baskets, 4 EX and 2 outlier groups
### Prepare model inputs
set.seed(99999)
# Set up data
rr_k_ctrl <- c(0.15,0.15, 0.15, 0.15, 0.15, 0.15)        # control response rate for each basket
rr_k_trt <- c(0.15,0.50, 0.70, 0.70, 0.70, 0.70)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(50, K)     # number of control participants per basket
N_k_trt <- rep(50, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

z_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  z_vec<-c(z_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=z_vec)

```

### Working model

Based on Equation *4* and the priors suggested in **TABLE 2** OF [Daniells et al, 2023, SIM](https://doi.org/10.1002/sim.9867). Using centered parameterizattion.

\$\$

\begin{aligned}
\mu_{\text{ex}} 
&\sim \text{Normal}(\text{logit}(0.15), 10) 
&& \text{(EX mean)} \\

\sigma_{\text{ex}} 
&\sim \text{Half-Normal}(0, 1) 
&& \text{(EX SD)} \\

\mu_{\text{nex}} 
&= -0.62 
&& \text{(NEX mean)} \\

\sigma_{\text{nex}} 
&= 4.4 
&& \text{(NEX SD)} \\

\alpha_{\text{k}} 
&\sim \text{Normal}(0, 5) 
&& \text{(Baseline log-odds, No borrowing)} \\

\mathbf{a}_{\text{Mix}} 
&= (2, 2) 
&& \text{(Dirichlet prior parameters)} \\[6pt]

\text{for } k = 1, \dots, K: \\

\mathbf{p}_k = (p_{k1}, p_{k2}) 
&\sim \text{Dirichlet}(\mathbf{a}_{\text{Mix}})
&& \text{(mixture weights)} \\

f_{\text{EX}}(\beta_k) 
&= \text{Normal}(\beta_k \mid \mu_{\text{ex}}, \sigma_{\text{ex}}) \\

f_{\text{NEX}}(\beta_k) 
&= \text{Normal}(\beta_k \mid \mu_{\text{nex}}, \sigma_{\text{nex}})
&& \text{(component densities)} \\

\log p(\beta_k \mid \mathbf{p}_k) 
&= \log \Big(
p_{k1} \cdot f_{\text{EX}}(\beta_k)
+ p_{k2} \cdot f_{\text{NEX}}(\beta_k)
\Big)
&& \text{(mixture prior)} \\[6pt]

\text{for } i = 1, \dots, N: \\

\text{logit}(p_i) 
&= \alpha_{k} + \beta_{k} z_i \\

y_i 
&\sim \text{Bernoulli}(p_i)
\end{aligned}

\$\$

```{r workingstanmod, include=TRUE,echo=TRUE,eval=TRUE}

#Working model
exnex_paper2 <- "data {
  int<lower=1> K;                         // number of baskets
  int<lower=1> N;                         // number of subjects
  array[N] int<lower=1, upper=K> k_vec;   // basket index
  array[N] int<lower=0, upper=1> z_vec;   // treatment indicator
  array[N] int<lower=0, upper=1> y;       // outcome

  vector<lower=0>[2] aMix;                // Dirichlet prior for mixture
}

transformed data {
  real nex_mu = -0.62;
  real nex_sigma = 4.4;             // convert variance → sd
}

parameters {
  array[K] simplex[2] pMix;        // mixture weights

  array[K] real alpha;    // baseline log-odds
  array[K] real beta;     // treatment effects

  // EX hyperparameters
  real ex_mu;
  real<lower=0> ex_sigma;
}

model {

  // EX priors (from paper)
  ex_mu ~ normal(logit(0.15), 10);

  // Half-normal(0,1) on sigma:
  ex_sigma ~ normal(0, 1);

  // Baseline (no borrowing)
  alpha ~ normal(0, 5);

  // EXNEX mixture prior on treatment effects
  for (k in 1:K) {
    pMix[k] ~ dirichlet(aMix);
    real lp_ex  = normal_lpdf(beta[k] | ex_mu, ex_sigma);
    real lp_nex = normal_lpdf(beta[k] | nex_mu, nex_sigma);

    target += log_mix(pMix[k][1], lp_ex, lp_nex);
  }

  // Likelihood
  for (i in 1:N) {
    y[i] ~ bernoulli_logit(alpha[k_vec[i]] + beta[k_vec[i]] * z_vec[i]);
  }
}

generated quantities {
  array[K] real p_control;     // response prob: control
  array[K] real p_treatment;   // response prob: treatment
  array[K] real diff;          // absolute difference
  array[K] real odds_ratio;    // odds ratio
  array[N] int y_rep;

  for (k in 1:K) {
    // Control arm probability
    p_control[k] = inv_logit(alpha[k]);

    // Treatment arm probability
    p_treatment[k] = inv_logit(alpha[k] + beta[k]);

    // Risk difference
    diff[k] = p_treatment[k] - p_control[k];

    // Odds ratio (optional but often useful)
    odds_ratio[k] = exp(beta[k]);
  }
  
   for (i in 1:N) {
    y_rep[i] = bernoulli_logit_rng(alpha[k_vec[i]] + beta[k_vec[i]] * z_vec[i]);
  }
}"

```

##Setup and run sampler

```{r datasets, echo=TRUE,results='hide',eval=TRUE}
# Create list with input values for Stan model
data_input_1 <- list(
  K = K,                # number of subgroups
  N = N,                # total sample size
  N_k = N_k,            # sample size per basket
  k_vec = k_vec,        # N x 1 vector of subgroup indicators
  z_vec = z_vec,        # N x 1 vector of active treatment arm indicators
  y = y,                 # N x 1 vector of binary responses
  aMix = c(2, 2)
  #nex_mu = qlogis(0.35),
  #nex_sigma = 4.4
  )



### Compile and fit model
options(mc.cores = parallel::detectCores())
# Compile model (only run the line below once for a simulation study - compilation not dependent on any
# simulation inputs as defined by scenarios or on simulated data)
stan_mod_1 <- stan_model(model_code = exnex_paper2)
# Fit model
nsamps <- 30000       # number of posterior samples (after removing burn-in) per chain
nburnin <- 1000       # number of burn-in samples to remove at beginning of each chain
nchains <- 4          # number of chains
#exnex_pars_1 <- c("p","pMix")     # parameters to sample
start_time <- Sys.time()
stan_fit_1 <- sampling(stan_mod_1, data = data_input_1, 
                       #pars = exnex_pars_1,
                       iter = nsamps + nburnin, warmup = nburnin, chains = nchains)

end_time <- Sys.time()
end_time - start_time
post_draws <- as.matrix(stan_fit_1)           # posterior samples of each parameter

post <- rstan::extract(stan_fit_1)

#mean(post$pMix)
#quantile(post$pMix[,1], c(0.025, 0.5, 0.975))
#colMeans(post$beta)
#mean(post$ex_mu)

cat("Observed response rates in the data")
with(thedata, tapply(y, list(basketID, Treatment), mean))


```

```{r stanresults,fig.width=10, fig.height=6, out.width="100%",eval=TRUE}
color_scheme_set("red")
plot_title <- ggtitle(
  "Treatment arm response rate/basket:  posterior distribution",
  "Median ± 79% credible interval highlighted (79 is arbitrary, like 95 — and a prime!)"
)
mcmc_areas(post_draws, 
           pars = c("p_treatment[1]", "p_treatment[2]", "p_treatment[3]", "p_treatment[4]",
                    "p_treatment[5]", "p_treatment[6]"), 
           prob = 0.79) + plot_title +
  theme(
    plot.subtitle = element_text(size = 10)
  )

```

```{r stanresults2,fig.width=10, fig.height=6, out.width="100%",eval=TRUE}
yrep <- rstan::extract(stan_fit_1)$y_rep

plot_title2 <- ggtitle("Posterior predictive distribution (1000 draws)")

ppc_dens_overlay(
  y = y,                 # your observed data
  yrep = yrep[1:100, ]    # first 50 posterior draws
) + plot_title2
```

```{r stanresults3,fig.width=10, fig.height=6, out.width="100%",eval=TRUE}
fit_df <- as.data.frame(stan_fit_1)

posterior2 <- extract(stan_fit_1, inc_warmup = TRUE, permuted = FALSE)

color_scheme_set("mix-blue-pink")
p <- mcmc_trace(posterior2,  
                pars = c("ex_mu", "ex_sigma"), 
                n_warmup = 300,
                facet_args = list(nrow = 2, labeller = label_parsed))
p + facet_text(size = 15)

```
